---
title: 'Data Visualization in R: Session 4'
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
date: "October 8, 2014"
---
```{r setup, echo=FALSE, message=FALSE}
require(knitr)
library(png)
library(grid)
#turn off mesages and warnings and make it so output isn't prefixed by anything,
#default is to put "##" in front of all output for some reason
#also set tidy to true so code is wrapped properly 
#opts_chunk$set(message=FALSE, warning=FALSE, comment = "", tidy = T)
opts_chunk$set(message=FALSE, warning=FALSE, comment = "", tidy=TRUE)
options(width = 60)
#options(rpubs.upload.method = "internal")

options(RCurlOptions = list(verbose = FALSE, capath = system.file("CurlSSL", "cacert.pem", package = "RCurl"), ssl.verifypeer = FALSE))

```

************

#Part 1 Installing and loading R packages
Now we can use the colors provided by R but we can also use an R package called `RColorBrewer` or `colorspace` to provide us with even more colors but first we have to learn how to download the packages which can be done using `install.packages()` and `library()` or we can use RStudio to download them 

Using `install.packages()` and `library()`
```{r, eval=FALSE}
#install package
install.packages("ggplot2")
#Once package is installed to actually use it you need to load it
library("ggplot2")
```  

Or we can use RStudio to manually install the packages  
In the bottom right corner of RStudio click the Packages tab and then the install icon  
```{r fig.width=5, fig.height=3, echo=FALSE}
grid.raster(readPNG("../images/installRPackage_1.png"))
```  

Type in ggplot2  

```{r fig.width=5, fig.height=2, echo=FALSE}
grid.raster(readPNG("../images/install_ggplot.png"))
```  

To Load click the check box next to ggplot2  

```{r fig.width=5, fig.height=3, echo=FALSE}
grid.raster(readPNG("../images/install_ggplot2.png"))
```  

You can see the output in the console window when doing this by hand and it actually is doing what we did before by calling `install.packages()` and `library()` 

```{r fig.width=5, fig.height=3, echo=FALSE}
grid.raster(readPNG("../images/install_ggplot3.png"))
```  


#Part 2: Reading in Data

Download new data set, frequency of several diseases in Boston since the 1880's [BosEpi.tab.txt](https://www.dropbox.com/s/mdz49u7hl751ypx/BosEpi.tab.txt?dl=0)

```{r, eval=FALSE}
bosEpi <- read.delim(file.choose())
```

```{r, echo=FALSE}
bosEpi <- read.delim("../dataSets/BosEpi.tab.txt")
worEpi <- read.delim("../dataSets/WorEpi.tab.txt")
sessionOne <- read.delim("../dataSets/Session_1.txt")
```

View data table 
```{r, eval=FALSE}
View(bosEpi)
```

Important columns of the data set  
**number** - Number of occurences  
**loc** - Location of disease  
**event** - Whether the number in **number** is number of cases or number of deaths from disease  
**disease** - Disease Name  
**Year** - Year of occurence   
**Month** - Month of occurrence   
**Day** - Day of occurrence   

# Part 3: Plotting using ggplot2

qplot() is a superior alternative to plot(). It is part of the "ggplot2" package and provides easy ways to produce publication-quality figures.  
For example:

```{r fig.width=10}
library(ggplot2)
qplot(Year, number, geom="histogram", stat="identity", fill=disease, position="dodge", data=bosEpi[bosEpi$event == "DEATHS",], xlim=c(1885, 1940))
```

# Part 4: step-by-step qplot()

We will now start with a few small examples of qplot().

First, let's generate some random data.

```{r}
randomData <- rnorm(10000) # normal
```

The default for qplot() is to produce a histogram.

```{r}
qplot(randomData)
```

First lets extract the data and sum it again like before   
```{r}
dipScarDeaths <- bosEpi[(bosEpi$disease == "DIPHTHERIA" | bosEpi$disease == "SCARLET FEVER") & bosEpi$event == "DEATHS",]

dipScarDeathsSum <- aggregate(number~disease*Year, dipScarDeaths, sum)
head(dipScarDeathsSum)

```

Now, this is how we can use qplot() to plot the number of deaths per year, using the  aggregated data.

```{r}
qplot(Year, number, data=dipScarDeathsSum)
```

In order to see which points correspond to which disease, we can color by disease.

```{r}
qplot(Year, number, data=dipScarDeathsSum, color=disease)
```

Also, we may not be interested in the diphtheria death outlier in 1980, so, let's change the scale of the x-axis.

```{r}
qplot(Year, number, data=dipScarDeathsSum, color=disease, xlim=c(1880, 1930))
```

What we see is that qplot() automatically chooses reasonable color and also produces a legend.

To turn this into a bar plot we can do the following.

```{r}
qplot(Year, number, data=dipScarDeathsSum, color=disease, xlim=c(1880, 1930), geom="bar", stat="identity", position="dodge", fill=disease, main = "Deaths by Diphtheria vs Deaths by Scarlet Fever", xlab = "Year", ylab = "Number of Deaths")
```

Last week we saw that we can use plot() to generate a similar figure. qplot() make this much simpler.

```{r}
plot(dipScarDeathsSum[dipScarDeathsSum$disease == "SCARLET FEVER",]$Year,
     dipScarDeathsSum[dipScarDeathsSum$disease == "SCARLET FEVER",]$number,
     type = "h", xlab = "Year", ylab = "Number of Deaths", 
     main = "Deaths by Diphtheria vs Deaths by Scarlet Fever",
     bty = "none", lwd = 2, xlim=c(1880, 1930), 
     ylim = range(dipScarDeathsSum$number), col= "red")  

points(dipScarDeathsSum[dipScarDeathsSum$disease == "DIPHTHERIA",]$Year + 0.5,
       dipScarDeathsSum[dipScarDeathsSum$disease == "DIPHTHERIA",]$number,
       type = "h", lwd = 2, col = "darkblue") 

legend("topright", legend = c("SCARLET FEVER", "DIPHTHERIA"),
       lwd = 2, col = c("red", "darkblue"))
```



